---
title: "Likelihood Assignment 2"
format: pdf
editor: visual
---

```{r}
library(readr)
coffeetime <- read_csv("coffeetime.csv")
View(coffeetime)
```

```{r}
library(cmdstanr)
```

## Q5

```{r}
mu <- mean(coffeetime$time)
sigma <- sd(coffeetime$time)

alpha_prior <- mu^{2}/ sigma^{2}
beta_pror <- mu/ sigma^2


```
I expect the average café waiting time to be around 4.4 minutes, with noticeable variation between cafés, perhaps with a standard deviation around 5 minutes. This corresponds to approximate Gamma hyperparameters α=0.78 and β=0.18. Therefore, I place exponential priors on α and β with these means.


```{r}
model <- cmdstan_model("Likelihood2.stan")
```

```{r}
stan_data <- list(
  N = length(coffeetime$cafe),
  n_cafes = 15,
  t = coffeetime$time,
  cafe = coffeetime$cafe
)
```

```{r}
fit <- model$sample(
  data = stan_data,
  chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  seed = 123
)
```

```{r}
sum <- fit$summary()

sum[] <- lapply(sum, function(x) {
  if(is.numeric(x)) round(x, 3) else x
})

kableExtra::kable(sum)
```


The MCMC diagnostics suggest that the chains have mixed well. All R-hat values are very close to 1, with the largest approximately 1.004, which indicates no obvious lack of convergence. The effective sample sizes are also large for both the bulk and tail of the posterior distributions, suggesting that the posterior summaries are reliable. Therefore, there is no clear evidence of MCMC convergence problems.

```{r}

hist(
  sum$mean[4:18],
  main = "Posterior Mean Waiting Times by Café",
  xlab = "Posterior Mean Waiting Time (minutes)",
  ylab = "Frequency",
  border = "white",
  col = "skyblue",
  breaks = 8
)

```
```{r}
library(posterior)

samples <- fit$draws(c("alpha", "beta"), format = "matrix")

mean_waiting_time <- samples[, "alpha"] / samples[, "beta"]


hist(
  mean_waiting_time,
  main = "Posterior Draws of the Mean Expected Waiting Time Across Cafés",
  xlab = "Mean expected waiting time across cafés (minutes)",
  ylab = "Frequency",
  breaks = 30,
  col = "skyblue",
  border = "white"
)
```



## Q7

```{r}

```
