---
title: "Likelihood Assignment 2"
format: pdf
editor: visual
---

```{r}
#| include: false
library(readr)
coffeetime <- read_csv("coffeetime.csv")
View(coffeetime)
```

```{r}
#| include: false
library(cmdstanr)
```

## Question 1

By conditional independence of the $T_i$ 's given $\lambda$,

$$
\pi(t \mid \lambda)
=
\prod_{i=1}^n \pi(t_i \mid \lambda_i).
$$

Also, since the $\lambda_i$ 's are iid given $\theta=(\alpha,\beta)$ ,

$$
\pi(\lambda \mid \theta)
=
\prod_{i=1}^n \pi(\lambda_i \mid \theta).
$$

Therefore,

$$
\pi(t \mid \lambda)\pi(\lambda \mid \theta)
=
\prod_{i=1}^n
\pi(t_i \mid \lambda_i)\pi(\lambda_i \mid \theta).
$$

Each term in the product depends only on one $\lambda_i$ , so the integral factorises:

$$
L(\theta)
=
\int
\prod_{i=1}^n
\pi(t_i \mid \lambda_i)\pi(\lambda_i \mid \theta)
\,d\lambda_1\cdots d\lambda_n
$$

$$
=
\prod_{i=1}^n
\int_0^\infty
\pi(t_i \mid \lambda_i)\pi(\lambda_i \mid \theta)
\,d\lambda_i.
$$

## Question 3

For repeated visits to café (i), the model is:

$$
T_{ij}\mid \lambda_i \sim \text{Exponential}(\lambda_i),
\qquad j=1,\ldots,v_i,\quad i=1,\ldots,n.
$$

The café-specific expected waiting times are modelled as

$$
\lambda_i \overset{iid}{\sim} \text{Gamma}(\alpha,\beta),
\qquad i=1,\ldots,n.
$$

Conditional on $\lambda_i$ , the waiting times from repeated visits to the same café are assumed independent.

## Question 4

Let $t_i=(t_{i1},\ldots,t_{iv_i})$ denote the waiting times observed at café (i).

The likelihood is obtained by marginalising over the unobserved $\lambda_i$ 's:

$$
L(\theta)
=
\pi(t\mid \theta)
=
\int \pi(t\mid \lambda)\pi(\lambda\mid \theta)\,d\lambda.
$$

Given $\lambda$ , observations from different cafés are independent, so

$$
\pi(t\mid \lambda)
=
\prod_{i=1}^n \pi(t_i\mid \lambda_i).
$$

Within café (i), repeated visits are independent conditional on $\lambda_i$ , hence

$$
\pi(t_i\mid \lambda_i)
=
\prod_{j=1}^{v_i}\pi(t_{ij}\mid \lambda_i).
$$

As before, the $\lambda_i$ 's are iid, therefore:

$$
\pi(\lambda\mid \theta)
=
\prod_{i=1}^n \pi(\lambda_i\mid \theta).
$$

Therefore,

$$
L(\theta)
=
\int
\prod_{i=1}^n
\left[
\left\{
\prod_{j=1}^{v_i}
\pi(t_{ij}\mid \lambda_i)
\right\}
\pi(\lambda_i\mid \theta)
\right]
d\lambda_1\cdots d\lambda_n.
$$

Since each bracketed term depends only on one $\lambda_i$ , the integral separates:

$$
L(\theta)
=
\prod_{i=1}^n
\int_0^\infty
\left[
\prod_{j=1}^{v_i}
\pi(t_{ij}\mid \lambda_i)
\right]
\pi(\lambda_i\mid \theta)
\,d\lambda_i.
$$

## Question 5

```{r}
mu <- mean(coffeetime$time)
sigma <- sd(coffeetime$time)

alpha_prior <- mu^{2}/ sigma^{2}
beta_pror <- mu/ sigma^2
```

It is expected that the average café waiting time will be around 4.4 minutes, with substantial variation between cafés, perhaps corresponding to a standard deviation of roughly 5 minutes. Matching these moments to a Gamma distribution gives approximate hyperparameter values of $\alpha \approx 0.78$ and $\beta \approx 0.18$.

Therefore the following exponential priors are placed on the Gamma hyperparameters:

$$
\alpha \sim \text{Exponential}\left(\frac{1}{0.78}\right),
\qquad
\beta \sim \text{Exponential}\left(\frac{1}{0.18}\right).
$$

These priors are relatively weakly informative, while still centering the model around plausible waiting times and between-café variability.


## Question 6

```{r}
#| echo: true
#| message: false
#| warning: false
random_effects_stan <- "
data {
  int<lower=0> N;
  vector<lower=0>[N] t;
  int<lower=1> n_cafes;
  array[N] int<lower=1, upper=n_cafes> cafe;
}

parameters {
  real<lower=0> alpha;
  real<lower=0> beta;
  vector<lower=0>[n_cafes] lambda;
}

model {
  alpha ~ exponential(1 / 0.7824712);
  beta  ~ exponential(1 / 0.1764025);

  lambda ~ gamma(alpha, beta);

  for (k in 1:N) {
    t[k] ~ exponential(1 / lambda[cafe[k]]);
  }
}
"

writeLines(random_effects_stan, "random_effects_model.stan")

model <- cmdstan_model("random_effects_model.stan")

```

```{r}
#| echo: true
#| message: false
#| warning: false
stan_data <- list(
  N = length(coffeetime$cafe),
  n_cafes = 15,
  t = coffeetime$time,
  cafe = coffeetime$cafe
)
```

```{r}
#| echo: true
#| message: false
#| warning: false
#| 
fit <- model$sample(
  data = stan_data,
  chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  seed = 123
)
```

```{r}
sum <- fit$summary()

sum[] <- lapply(sum, function(x) {
  if(is.numeric(x)) round(x, 3) else x
})

kableExtra::kable(sum)
```

The MCMC diagnostics suggest that the random effects model has fitted successfully. The $\hat{R}$ values are close to 1, which suggests that the chains have mixed well and are sampling from the same posterior distribution. The effective sample sizes are also reasonably large, meaning that the posterior summaries should be reliable. There is therefore no obvious evidence of non-convergence or poor mixing.

The posterior distribution of the café-specific expected waiting times shows that there is variation between cafés. This supports the use of a random effects model, since the model allows each café to have its own expected waiting time while still assuming that the cafés come from a common population distribution.

```{r}
#| echo: true
#| message: false
#| warning: false

hist(
  sum$mean[4:18],
  main = "Posterior Mean Waiting Times by Café",
  xlab = "Posterior Mean Waiting Time (minutes)",
  ylab = "Frequency",
  border = "white",
  col = "skyblue",
  breaks = 8
)

```

```{r}
#| echo: true
#| message: false
#| warning: false

samples <- fit$draws(c("alpha", "beta"), format = "matrix")

mean_waiting_time <- samples[, "alpha"] / samples[, "beta"]


hist(
  mean_waiting_time,
  main = "Posterior Draws of the Mean Expected Waiting Time Across Cafés",
  xlab = "Mean expected waiting time across cafés (minutes)",
  ylab = "Frequency",
  breaks = 30,
  col = "skyblue",
  border = "white"
)
```

## Question 7

For the fixed effects model, we still allow the expected waiting time to differ by café, but we do not assume that the café-specific parameters are drawn from a shared Gamma distribution. Thus, there is no pooling of information between cafés.

Let $T_{ij}$ be the waiting time for visit j to café i. I specify$T_{ij}\mid \lambda_i \sim \text{Exponential}(\lambda_i), \qquad j=1,\ldots,v_i,\quad i=1,\ldots,n$

so that $(T_{ij}\mid \lambda_i)=\lambda_i$.

For the café-specific expected waiting times, independent priors are used:

$\lambda_i \overset{ind}{\sim} \text{Exponential}(4.4) \qquad i=1,\ldots,n$

I chose 4.4 because in the random effects model, I expected the average café waiting time to be around 4.4 minutes. However, unlike the random effects model, this fixed effects model does not estimate shared parameters such as $\alpha$ and $\beta$. Therefore the cafés are not linked through a common estimated population distribution and the posterior estimate of each $\lambda_i$ is informed only by the data from only café i and its own prior.

## Q8

```{r}
#| echo: true
#| message: false
#| warning: false


fixed_effects_stan <- "
data {
  int<lower=0> N;
  vector<lower=0>[N] t;
  int<lower=1> n_cafes;
  array[N] int<lower=1, upper=n_cafes> cafe;
}

parameters {
  vector<lower=0>[n_cafes] lambda;
}

model {
  lambda ~ exponential(1 / 4.4);

  for (k in 1:N) {
    t[k] ~ exponential(1 / lambda[cafe[k]]);
  }
}
"

writeLines(fixed_effects_stan, "fixed_effects_model.stan")

fixed_mod <- cmdstan_model("fixed_effects_model.stan")
```


```{r}
#| echo: true
#| message: false
#| warning: false

library(tidyverse)

# Fit model
fixed_fit <- fixed_mod$sample(
  data = stan_data,
  chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  seed = 123
)

# Check summary
fixed_fit$summary()
```

Checking MCMC for Q8

```{r}
#| echo: true
#| message: false
#| warning: false


sum_fixed <- fixed_fit$summary()

sum_fixed[] <- lapply(sum_fixed, function(x) {
  if(is.numeric(x)) round(x, 3) else x
})

kableExtra::kable(sum_fixed)
```

The MCMC diagnostics suggest that the model converged well. All ($\hat{R}$) values are extremely close to 1, indicating good mixing and convergence across chains. In addition, both the bulk and tail effective sample sizes are large for all parameters, suggesting low autocorrelation and reliable estimation of posterior summaries and credible intervals. Therefore, there is no evidence of concerning convergence or sampling issues in the fitted model.


## Q9


```{r}

#| echo: true
#| message: false
#| warning: false

# Random effects sum
rand_sum <- fit$summary()

rand_cafe <- rand_sum |>
  filter(str_detect(variable, "^lambda\\[")) |>
  mutate(
    cafe = as.numeric(str_extract(variable, "\\d+")),
    model = "Random effects"
  ) |>
  select(cafe,model, median,q5,q95)

# Fixed efects sum
fixed_sum <- fixed_fit$summary()

fixed_cafe <- fixed_sum |>
  filter(str_detect(variable, "^lambda\\[")) |>
  mutate(
    cafe = as.numeric(str_extract(variable, "\\d+")),
    model = "Fixed effects"
  ) |>
  select(cafe,model,median,q5,q95)

# Posterior median of overall mean expect waiting time
rand_draws <- fit$draws(c("alpha", "beta"), format = "df")

mu_draws <- rand_draws$alpha / rand_draws$beta
mu_median <- median(mu_draws)



plot_data <- bind_rows(rand_cafe, fixed_cafe)

ggplot(plot_data, aes(x = factor(cafe), y = median, colour = model)) +
  geom_hline(
    yintercept = mu_median,
    linetype = "dashed",
    colour = "black"
  ) +
  geom_pointrange(
    aes(ymin = q5, ymax = q95),
    position = position_dodge(width = 0.5)
  ) +
  labs(
    x = "Café",
    y = "Expected waiting time",
    colour = "Model",
    title = "Comparison of fixed effects and random effects estimates",
    subtitle = "Posterior medians and 90% credible intervals"
  ) +
  theme_minimal()
```

The plot shows the general difference between the fixed effects and random effects models. The fixed effects model estimates each café separately, so its café-specific estimates are based mainly on the data from that café alone. As a result, the estimates can vary more strongly between cafés, especially for cafés with less data.

The random effects model partially pools information across cafés. This means that café-specific estimates are pulled towards the overall mean expected waiting time. Therefore, cafés with unusually high or low fixed effects estimates tend to have less extreme estimates under the random effects model. The random effects credible intervals may also be narrower or more stable because the model borrows strength across cafés.

The horizontal dashed line represents the posterior median of the overall mean expected waiting time from the random effects model. The random effects café estimates are generally shrunk towards this line, while the fixed effects estimates are allowed to move more freely away from it. This pattern is the main evidence of partial pooling in the random effects model.


## Q10

In standard asymptotic theory, standard errors of parameter estimates typically shrink at a rate of $\frac{1}{\sqrt{n}}$ as the sample size increases. In this context, this means that the standard errors will initially decrease as we gain more information.

However, the mean waiting time parameter is designed to describe the whole population of cafés, and continuously sampling from only one café does not provide sufficient information about the population as a whole. While repeated visits give increasingly precise information about Café Statistique itself, they provide little additional information about between-café variability. Therefore, the standard error for the population mean waiting time will decrease initially, but will not shrink to zero as n becomes very large, and will instead plateau.
