---
title: "Likelihood Assignment 2"
format: pdf
editor: visual
---

```{r}
library(readr)
coffeetime <- read_csv("coffeetime.csv")
View(coffeetime)
```

```{r}
library(cmdstanr)
```

## Question 1

By conditional independence of the $T_i$ 's given $\lambda$,

$$
\pi(t \mid \lambda)
=
\prod_{i=1}^n \pi(t_i \mid \lambda_i).
$$

Also, since the $\lambda_i$ 's are iid given $\theta=(\alpha,\beta)$ ,

$$
\pi(\lambda \mid \theta)
=
\prod_{i=1}^n \pi(\lambda_i \mid \theta).
$$

Therefore,

$$
\pi(t \mid \lambda)\pi(\lambda \mid \theta)
=
\prod_{i=1}^n
\pi(t_i \mid \lambda_i)\pi(\lambda_i \mid \theta).
$$

Each term in the product depends only on one $\lambda_i$ , so the integral factorises:

$$
L(\theta)
=
\int
\prod_{i=1}^n
\pi(t_i \mid \lambda_i)\pi(\lambda_i \mid \theta)
\,d\lambda_1\cdots d\lambda_n
$$

$$
=
\prod_{i=1}^n
\int_0^\infty
\pi(t_i \mid \lambda_i)\pi(\lambda_i \mid \theta)
\,d\lambda_i.
$$

## Question 3

For repeated visits to café (i), the model is:

$$
T_{ij}\mid \lambda_i \sim \text{Exponential}(\lambda_i),
\qquad j=1,\ldots,v_i,\quad i=1,\ldots,n.
$$

The café-specific expected waiting times are modelled as

$$
\lambda_i \overset{iid}{\sim} \text{Gamma}(\alpha,\beta),
\qquad i=1,\ldots,n.
$$

Conditional on $\lambda_i$ , the waiting times from repeated visits to the same café are assumed independent.

## Question 4

Let $t_i=(t_{i1},\ldots,t_{iv_i})$ denote the waiting times observed at café (i).

The likelihood is obtained by marginalising over the unobserved $\lambda_i$ 's:

$$
L(\theta)
=
\pi(t\mid \theta)
=
\int \pi(t\mid \lambda)\pi(\lambda\mid \theta)\,d\lambda.
$$

Given $\lambda$ , observations from different cafés are independent, so

$$
\pi(t\mid \lambda)
=
\prod_{i=1}^n \pi(t_i\mid \lambda_i).
$$

Within café (i), repeated visits are independent conditional on $\lambda_i$ , hence

$$
\pi(t_i\mid \lambda_i)
=
\prod_{j=1}^{v_i}\pi(t_{ij}\mid \lambda_i).
$$

As before, the $\lambda_i$ 's are iid, therefore:

$$
\pi(\lambda\mid \theta)
=
\prod_{i=1}^n \pi(\lambda_i\mid \theta).
$$

Therefore,

$$
L(\theta)
=
\int
\prod_{i=1}^n
\left[
\left\{
\prod_{j=1}^{v_i}
\pi(t_{ij}\mid \lambda_i)
\right\}
\pi(\lambda_i\mid \theta)
\right]
d\lambda_1\cdots d\lambda_n.
$$

Since each bracketed term depends only on one $\lambda_i$ , the integral separates:

$$
L(\theta)
=
\prod_{i=1}^n
\int_0^\infty
\left[
\prod_{j=1}^{v_i}
\pi(t_{ij}\mid \lambda_i)
\right]
\pi(\lambda_i\mid \theta)
\,d\lambda_i.
$$

## Q5

```{r}
mu <- mean(coffeetime$time)
sigma <- sd(coffeetime$time)

alpha_prior <- mu^{2}/ sigma^{2}
beta_pror <- mu/ sigma^2


```

I expect the average café waiting time to be around 4.4 minutes, with noticeable variation between cafés, perhaps with a standard deviation around 5 minutes. This corresponds to approximate Gamma hyperparameters α=0.78 and β=0.18. Therefore, I place exponential priors on α and β with these means.

```{r}
model <- cmdstan_model("Likelihood2.stan")
```

```{r}
stan_data <- list(
  N = length(coffeetime$cafe),
  n_cafes = 15,
  t = coffeetime$time,
  cafe = coffeetime$cafe
)
```

```{r}
fit <- model$sample(
  data = stan_data,
  chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  seed = 123
)
```

```{r}
sum <- fit$summary()

sum[] <- lapply(sum, function(x) {
  if(is.numeric(x)) round(x, 3) else x
})

kableExtra::kable(sum)
```

The MCMC diagnostics suggest that the chains have mixed well. All R-hat values are very close to 1, with the largest approximately 1.004, which indicates no obvious lack of convergence. The effective sample sizes are also large for both the bulk and tail of the posterior distributions, suggesting that the posterior summaries are reliable. Therefore, there is no clear evidence of MCMC convergence problems.

```{r}

hist(
  sum$mean[4:18],
  main = "Posterior Mean Waiting Times by Café",
  xlab = "Posterior Mean Waiting Time (minutes)",
  ylab = "Frequency",
  border = "white",
  col = "skyblue",
  breaks = 8
)

```

```{r}
library(posterior)

samples <- fit$draws(c("alpha", "beta"), format = "matrix")

mean_waiting_time <- samples[, "alpha"] / samples[, "beta"]


hist(
  mean_waiting_time,
  main = "Posterior Draws of the Mean Expected Waiting Time Across Cafés",
  xlab = "Mean expected waiting time across cafés (minutes)",
  ylab = "Frequency",
  breaks = 30,
  col = "skyblue",
  border = "white"
)
```

## Q7

For the fixed effects model, we still allow the expected waiting time to differ by café, but we do not assume that the café-specific parameters are drawn from a shared Gamma distribution. Thus, there is no pooling of information between cafés.

Let $T_{ij}$ be the waiting time for visit j to café i. I specify$T_{ij}\mid \lambda_i \sim \text{Exponential}(\lambda_i), \qquad j=1,\ldots,v_i,\quad i=1,\ldots,n$

so that $(T_{ij}\mid \lambda_i)=\lambda_i$.

For the café-specific expected waiting times, independent priors are used:

$\lambda_i \overset{ind}{\sim} \text{Exponential}(4.4) \qquad i=1,\ldots,n$

I chose 4.4 because in the random effects model, I expected the average café waiting time to be around 4.4 minutes. However, unlike the random effects model, this fixed effects model does not estimate shared parameters such as $\alpha$ and $\beta$. Therefore the cafés are not linked through a common estimated population distribution and the posterior estimate of each $\lambda_i$ is informed only by the data from only café i and its own prior.

## Q8

```{r}
library(tidyverse)

# Compile fixed effects model
fixed_mod <- cmdstan_model("Q8Stan.stan")

# Fit model
fixed_fit <- fixed_mod$sample(
  data = stan_data,
  chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  seed = 123
)

# Check summary
fixed_fit$summary()
```

Checking MCMC for Q8

```{r}
sum_fixed <- fixed_fit$summary()

sum_fixed[] <- lapply(sum_fixed, function(x) {
  if(is.numeric(x)) round(x, 3) else x
})

kableExtra::kable(sum_fixed)
```

The MCMC diagnostics suggest that the model converged well. All (\hat{R}) values are extremely close to 1, indicating good mixing and convergence across chains. In addition, both the bulk and tail effective sample sizes are large for all parameters, suggesting low autocorrelation and reliable estimation of posterior summaries and credible intervals. Therefore, there is no evidence of concerning convergence or sampling issues in the fitted model.

## Q10

In standard asymptotic theory, standard errors of parameter estimates typically shrink at a rate of $\frac{1}{\sqrt{n}}$ as the sample size increases. In this context, this means that the standard errors will initially decrease as we gain more information.

However, the mean waiting time parameter is designed to describe the whole population of cafés, and continuously sampling from only one café does not provide sufficient information about the population as a whole. While repeated visits give increasingly precise information about Café Statistique itself, they provide little additional information about between-café variability. Therefore, the standard error for the population mean waiting time will decrease initially, but will not shrink to zero as n becomes very large, and will instead plateau.
